Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-05 manuscript/manuscript.tex
Finish definitions section.

\documentclass[11pt, letterpaper]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage[backend=biber, style=authoryear, autocite=inline, isbn=false, doi=false, url=false, eprint=false]{biblatex}
% Fix biblatex behavior that writes "In: " before journal name.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\E}[1]{\left< #1 \right>}

@@

\section*{Definitions and non-recombining theory}

\subsection*{The SFS and 2-SFS}
Consider a sample of $n$ chromosomes taking from a population of size $N$ with mutation rate $\mu$ per site per generation.
Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : i \in \{1,\ldots,n-1\} \right\}$, where $\xi_i$ is the fraction of sites at which there is a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}: i \in \{1, \ldots, \floor{n/2} \right\}$. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.

The SFS and folded SFS are single-site statistics.
As such, they can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
\begin{align*}
    \left\{ \xi_{ij}(d) : i \in \{1, \ldots, n-1\}, j \in \{i, \ldots, n-1\} \right\},
\end{align*}
as the fraction of pairs of sites separated by $d$ bases with a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
By analogy to the folded SFS, we define the folded 2-SFS, as
\begin{align*}
    \eta_{ij}(d) &= \xi_{ij}(d) \\
                 &+ \delta_{n-2i} \xi_{(n-i),j}(d) \\
                 &+ \delta_{n-2j} \xi_{i,(n-j)}(d) \\
                 &+ \delta_{n-2i} \delta_{n-2j} \xi_{(n-i),(n-j)}(d).
\end{align*}
In this section, we will be considering non-recombining sites and so will suppress the distance $d$.

In the limit of low per-site mutation rate, the expected SFS and 2-SFS are related to moments of the branch length distribution through,
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
This result implies a negative association between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
(There are positive associations between mutations at complimentary frequencies induced by trees whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

Later, \cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003} and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \cite{NeherHallatscheck2013}.
Like Fu, Birkner et al. were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}, and thus mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

Our goal is to demonstrate that this distinction in the two-site frequency spectrum between multiple mergers and Kingman coalescent
(1) applies to Kingman coalescents with time-varying coalescent rates,
(2) is robust to recombination between the sites,
(3) is also a feature of forward-time models with selection,
and (4) can form the basis for a model-checking procedure for demographic inference methods.
To this end, we will make use of the fact that the 2-SFS may be seen as a joint probability distribution to define a useful data transformation.

\subsection*{Point-wise mutual information}
\begin{itemize}
    \item the sfs (with zeros) is a probability distribution
    \item the 2-sfs (with zeros) is a joint probability distribution
    \item Mutual information is a functional on joint pmfs that measures the association between two variables $\text{MI}[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)}$
    \item The contributions to MI from each variable are ``pointwise mutual information'' $\text{PMI}(x,y) = \log \frac{p(x,y)}{p(x)p(y)}$.
    \item Thus, we calculate PMI of the expected (folded) 2-SFS as $\text{PMI}(i,j) = \log \left( \frac{\E{\eta_{ij}}}{\E{\eta_i} \E{\eta_j}} \right)$
    \item The advantages of folding are (1) don't need ancestral state and (2) all elements are negative for standard Kingman.
    \item also, note that it doesn't depend on either the mutation rate or the average pairwise coalescent time $T_2$.
\end{itemize}

\subsection*{Population growth vs. the beta coalescent}

@@

\section*{Simulations with recombination}
[Coalescent simulations with recombination (it still works).]

\section*{Linked selective sweeps}
[Forward-time simulations of linked sweeps (still works).]

\section*{Application to \textit{Drosophila melanogaster}}

>>>>

\documentclass[11pt, letterpaper]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
% Allow inline lists
\usepackage[inline]{enumitem}
\usepackage[backend=biber, style=authoryear, autocite=inline, isbn=false, doi=false, url=false, eprint=false]{biblatex}
% Fix biblatex behavior that writes "In: " before journal name.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\E}[1]{\left< #1 \right>}
\newcommand{\eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Eq}[1]{Equation~(\ref{#1})}
\newcommand{\eqs}[2]{Eq.~(\ref{#1})~and~(\ref{#2})}
\DeclareMathOperator{\pmi}{pmi}
\DeclareMathOperator{\fpmi}{fpmi}
\DeclareMathOperator{\hilopmi}{hilopmi}

@@

\section*{Definitions and previously known results}

\subsection*{The SFS and 2-SFS}
Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.

The SFS and folded SFS are single-site statistics.
As such, they can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
$\left\{ \xi_{ij}(d) : 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
By analogy to the folded SFS, we define the folded 2-SFS, as
\begin{align*}
    \eta_{ij}(d) &= \xi_{ij}(d) \\
                 &+ (1-\delta_{i,n-i}) \xi_{n-i,j}(d) \\
                 &+ (1-\delta_{j,n-j}) \xi_{i,n-j}(d) \\
                 &+ (1-\delta_{i,n-i}) (1-\delta_{j,n-j}) \xi_{n-i,n-j}(d).
\end{align*}
In this section, we will be considering non-recombining sites and so will suppress the distance $d$.

In the limit of low per-site mutation rate ($\mu\to0$), all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \label{eq:expected_sfs} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
    \label{eq:expected_2sfs}
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
(There are positive correlations between mutations at complimentary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in rapidly adapting populations \autocite{NeherHallatscheck2013}.
Like Fu, Birkner et al.\ were primarily concerned with computing the sample variance of SFS-based summary statistics such as Tajima's D \autocite{Tajima1989}
As a result, they were mostly interested in the diagonal terms of the SFS covariance matrix, which dominate that calculation.
However, Figures 5 and 6 of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \neq i, n-i$.
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

Our goal is to demonstrate that this distinction in the two-site frequency spectrum between the multiple merger and Kingman coalescents:
\begin{enumerate*}
    \item applies to Kingman coalescents with time-varying coalescent rates,
    \item is robust to recombination between the sites,
    \item is also a feature of forward-time models with selection, and
    \item can form the basis of a model-checking procedure for demographic inference methods.
\end{enumerate*}
To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following useful transformation.

\subsection*{Frequency pointwise mutual information}

Given a coalescent model, the minor allele count at an arbitrary site is a random variable with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
We are interested in quantifying the dependence between the minor allele counts at these sites, particularly for $i\neq j$.

A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
The mutual information between random variables $X$ and $Y$ is a functional of their joint probability mass function, $p(x,y)$, defined as
\begin{align}
    I[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)},
    \label{eq:mutual_information}
\end{align}
where, with slight abuse of notation, $p(x) = \sum_y p(x,y)$ and vice versa.
Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of independent random variables with the same marginal distributions \autocite{CoverThomas1991}.
Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.

We would like to measure the association between alleles at particular frequencies, rather than the overall dependence between allele frequencies at nearby sites (which would be a measure of linkage disequilibrium).
\Eq{eq:mutual_information} shows that the $I$ is the expectation, relative to $p(x,y)$, of a quantity known as the \textit{pointwise mutual information}, where
\begin{align}
    \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)} = \log \frac{p(y|x)}{p(y)}
\end{align}
\autocite{ChurchHanks1990}.
Pointwise mutual information measures the change in the probability that $X=x$ given knowledge that $Y=y$.
When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.

As we have shown above, the folded 2-SFS is a joint probability distribution over minor allele counts.
We may thus define the \textit{frequency pointwise mutual information} as
\begin{align}
    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
\end{align}
This transformation of the 2-SFS has several useful properties.
First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
Moreover, the results of \cite{Fu1995} show that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.

The second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the SFS.
Population-size variation distorts the SFS by changing the distribution of waiting times.
However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous version.
On the other hand, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
Our hope is that the 2-SFS contains additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.

\subsection*{Coarse-grained frequencies}

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
This suggests a coarse-grained SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Coarse-graining allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.

We can also compute the pointwise mutual information in this coarse-grained distribution as
\begin{align}
    \hilopmi(d; i_c) = \log \frac
                                {\E{\eta_{\text{hi,lo}}(d; i_c)}}
                                {\E{\eta_{\text{lo}}(i_c)}  \E{\eta_{\text{hi}}(i_c)}}.
\end{align}
One could similarly calculate five other pointwise mutual information statistics from the coarse-grained 2-SFS (e.g the PMI between monomorphic and high minor allele count sites), but we will not use these statistics here.

\section*{Results \label{sec:results}}

\subsection*{Population growth vs. the beta coalescent in non-recombining loci}

@@

\subsection*{Pointwise mutual information between recombining sites}
[Coalescent simulations with recombination (it still works).]

\subsection*{Linked selective sweeps}
[Forward-time simulations of linked sweeps (still works).]

\subsection*{Application to \textit{Drosophila melanogaster}}
